### 2011-06-09 draft2.tex
pdf update

\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
\newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
\newcommand{\cf}[1]{\mbox{CF}(#1)}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

@@

\subsection{$\lambda$-lifted haskell subset}
\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n & (Expression)\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n = e \mid f~x_1 \dots x_n = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = D_1 ; D_2 ; \dots ; D_n & (Data~type~definition)\\ %TODO make type clearer
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

In the calculus, every definition is toplevel ($\lambda$-lifted) and there isn't any nested case.

@@

@@

\section{Translation}
We define several translations: $\etrans{},\dtrans{},\ttrans{},\strans{},\trans{}$.
\begin{center}
\begin{array}{rclr}
  \etrans{} &::& Expression \to Term\\
  \dtrans{} &::& Definition \to FOF\\
  \ttrans{} &::& Data~type  \to \{ FOF \}\\
  \strans{} &::& Expression \to Contract \to FOF\\
  \trans {} &::& Program    \to \{ FOF \}\\
\end{array}
\end{center}

\subsection{$\etrans{}$}
$\etrans{e}$ is a term, the translation is really straightforward.

\begin{eqnarray}
\etrans{x} &=& x\\
\etrans{f} &=& f\\
\etrans{e_1~e_2} &=& app(e_1,e_2)\\
\etrans{D(e_1,\dots,e_n)} &=& D(\etrans{e_1},\dots,\etrans{e_n})\\
\etrans{\bad} &=& \bad\\
\etrans{n} &=& n\\
\end{eqnarray}

As usual, we use left associativity.

\subsection{$\ttrans{}$}
$\ttrans{T}$ is a set of first-order formulae which we break down in four parts:
$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3 \cup S_4$.

@@

Finally, we have to say that none of the $D_i$ is unreachable:
$$S_4 := \{ D_i(x_1,\dots,x_n) = \unr \to x_1 = \unr \lor \dots \lor x_n = \unr \mid 1 \leq i \leq n \}$$
NB: We first started with $S_4 := \{ D_i \neq \unr \mid 1 \leq i \leq n \}$, but we couldn't prove that the (add x y), defined with peano integers, satisfied ($x=0 \to y=0 \to add~x~y=0$). It's also bizarre to state that the function was $\unr$ but not its application.

\subsection{$\dtrans{}$}
$\dtrans{d}$ is a first-order formula.

\begin{eqnarray}
\dtrans{f~\overline{x} = e} &=& \forall x_1 \dots x_n.\etrans{f~x_1 \dots x_n} = \etrans{e}\\
\dtrans{f~\overline{x} = \mbox{case } e \mbox{ of } [D_i(\overline{z}) \mapsto e_i]} &=& \forall x_1 \dots x_n.(\bigwedge_i (\forall \overline{z}~\etrans{e} = \etrans{D_i(\overline{z})} \to \etrans{f~\overline{x}} = \etrans{e_i})\\
&& \land \etrans{e} = \bad \to \etrans{f~x_1 \dots x_n} = \bad)\\ 
&& \land ((\bigwedge_i e \neq D_i(sel_{1,D_i}(e),\dots,sel_{n_i,D_i}(e)) \land \etrans{e} \neq \bad)\\
&& ~~\to \etrans{f~x_1 \dots x_n} = \unr)
\end{eqnarray}

@@

\subsection{$\strans{}$}
$\strans{e \in c}$ is a first-order formula.
\begin{eqnarray}
\strans{e \in \any} &=& true\\
\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etrans{e}) \land \etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)\\
\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
\end{eqnarray}
$False$ is a data constructor here.

Remark: we follow the semantics of the POPL paper but it's a bit restrictive.
e.g. in equation 14 we could use the alternate semantics (namely B1 in the POPL paper) : 
$$\strans{e \in \{x \mid u \}} = e = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False)$$

@@

We then want to check (with equinox) that:
 $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c} \} \models \bigwedge_{c \in C_f} \strans{f \in c}$$

Which we rewrite as:
 $$\trans{C} \cup \{ \dtrans{f~x_1 \dots \x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c},  \bigvee_{c \in C_f} \lnot \strans{f \in c} \} \models \bot$$

>>>>

\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\etranst}[1]{\mathcal{E}_t \llbracket #1 \rrbracket}
\newcommand{\etransf}[1]{\mathcal{E}_{f} \llbracket #1 \rrbracket}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
\newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
\newcommand{\cf}[1]{\mbox{CF}(#1)}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

@@

\subsection{$\mathcal{H}$: Haskell subset}
Our source language $\mathcal{H}$ is defined below. Note that $\lambda$ abstractions and case expressions can appear anywhere in the code.

assert $(e,c)$ in $e'$ means that if $e$ does not satisfy the contract $c$ then the compiler should return $\BAD$ and otherwise execute $e'$. % TODO return bad? Not really...

$e$ `satisfies` $c$ is a boolean expression which is True iff the expression $e$ satisfies the contract $c$ in the current context. % TODO we talk about cotnract satisfaction in the first paragraph!

\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid \lambda x.~e \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \mbox{assert } (e,c) \mbox{ in } e  & (Expression)\\
  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] &\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n ::: c = e \mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

\subsection{$\mathcal{H'}$: $\lambda$-lifted haskell subset}
This language is as expressive as $\mathcal{H}$ but contains fewer constructs. Compared to $\mathcal{H}$ the restrictions are:
\begin{itemize}
  \item No $\lambda$, ie every $\lambda$ is lifted to a function definition at the toplevel
  \item No case expressions which are lifted out to the toplevel (ie pattern matching is only possible for a function declaration)
  \item No $assert$ which are syntactic sugar for case expressions
\end{itemize} % TODO describe preprocessing.
From now on, expressions will always refer to $\mathcal{H'}$ expressions.

\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c & (Expression)\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n ::: c = e\mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

@@

\subsection{From $\mathcal{H}$ to $\mathcal{H'}$}
\subsubsection{Step 1: $assert$}
We translate every expression $e = \mbox{assert } (e_c,c) \mbox{ in } e'$ in $\mbox{case } e_c `satisfies` c \mbox{ of } True -> e' | False -> \bad$.
\subsubsection{Step 2: $\lambda-lifting$}
GHC's API can do that.
\subsubsection{Step 3: $case$ lifting}
...

@@

\section{Translation}
We define several translations: $\etrans{},\dtrans{},\ttrans{},\strans{},\trans{}$.
\begin{center}
\begin{array}{rclr}
  \etrans{} &::& Expression \to (Term,FOF)\\
  \dtrans{} &::& Definition \to \{ FOF \}\\
  \ttrans{} &::& Data~type  \to \{ FOF \}\\
  \strans{} &::& Expression \to Contract \to \{ FOF \}\\
  \trans {} &::& Program    \to \{ FOF \}\\
\end{array}
\end{center}

\subsection{$\etrans{}$}
$\etrans{e}$ is a pair (term,formula), the only tricky translations is
for satisfies.  We write $\etranst{}$ and $\etransf{}$ the projection
of $\etrans{}$ on the first and second component.

\begin{eqnarray}
\etrans{x} &=& (x,\emptyset)\\
\etrans{f} &=& (f,\emptyset)\\
\etrans{e_1~e_2} &=& (app(\etranst{e_1},\etranst{e_2}),\etransf{e_1} \cup \etransf{e_2})\\
\etrans{D(e_1,\dots,e_n)} &=& (D(\etranst{e_1},\dots,\etranst{e_n}),\cup_{1 \leq i \leq n} \etransf{e_i})\\
\etrans{\bad} &=& (\bad,\emptyset)\\
\etrans{n} &=& (n,\emptyset)\\
\etrans{e \mbox{ `satisfies` } c} &=& (\etranst{e},\etransf{e} \cup \forall x. \strans{x \in c} \leftrightarrow x~`satisfies`~c = True)\\
\end{eqnarray}

TODO: take car of the FV in c for satisfies.
As usual, we use left associativity.

\subsection{$\ttrans{}$}
$\ttrans{T}$ is a set of first-order formulae which we break down in five parts:
$\ttrans{\mbox{data } T = D_1 ; \dots ; D_n} = S_1 \cup S_2 \cup S_3 \cup S_4 \cup S_5$.

@@

We also have to say that none of the $D_i$ is unreachable:
$$S_4 := \{ D_i(x_1,\dots,x_n) = \unr \to x_1 = \unr \lor \dots \lor x_n = \unr \mid 1 \leq i \leq n \}$$
NB: We first started with $S_4 := \{ D_i \neq \unr \mid 1 \leq i \leq n \}$, but we couldn't prove that the (add x y), defined with peano integers, satisfied ($x=0 \to y=0 \to add~x~y=0$). It's also bizarre to state that the function was $\unr$ but not its application.

Finally, we give contracts to each data constructors. It is the contract $c$ specified by the user. If no contract is given for a specific data constructor, then we assume that the contract is $\ok \to \dots \to \ok$.
We write the contract $c$ as $c_1 \to \dots \to c_{n_i}}$
$$S_5 := \{ \forall x_1,\dots,x_{n_i} \strans{D_i(x_1,\dots,x_{n_i}) \in c} \leftrightarrow \bigwedge_{1 \leq i \leq n_i} \strans{x_i \in c_i} \mid 1 \leq i \leq n \}$$

\subsection{$\dtrans{}$}
$\dtrans{d}$ is a set first-order formula.

\begin{eqnarray}
\dtrans{f~\overline{x} = e} &=& \forall x_1 \dots x_n.\etranst{f~x_1 \dots x_n} = \etranst{e} \cup \etransf{e}\\
\dtrans{f~\overline{x} = \mbox{case } e \mbox{ of } [D_i(\overline{z}) \mapsto e_i]} &=& \forall x_1 \dots x_n.(\bigwedge_i (\forall \overline{z}~\etranst{e} = \etranst{D_i(\overline{z})} \to \etranst{f~\overline{x}} = \etranst{e_i})\\
&& \land \etranst{e} = \bad \to \etranst{f~x_1 \dots x_n} = \bad)\\ 
&& \land ((\bigwedge_i e \neq D_i(sel_{1,D_i}(e),\dots,sel_{n_i,D_i}(e)) \land \etranst{e} \neq \bad)\\
&& ~~\to \etranst{f~x_1 \dots x_n} = \unr)\\
&& \cup \etransf{e}
\end{eqnarray}

Note that in those equations, there is no $\etransf{f~x_1 \dots x_n}$ because it is always the empty set.

@@

\subsection{$\strans{}$}
$\strans{e \in c}$ is a set of first-order formulae.
\begin{eqnarray}
\strans{e \in \any} &=& true\\
\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etranst{e}) \land \etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
&& \cup \etransf{e} \cup \etransf{u[e/x]}\\
\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
\end{eqnarray}
$False$ is a data constructor here.

Remark: we follow the semantics of the POPL paper but it's a bit restrictive.
e.g. in equation 14 we could use the alternate semantics (namely B1 in the POPL paper) : 
$$\strans{e \in \{x \mid u \}} = e = \unr \lor (\etrans{u[e/x]} \neq \bad \land \etrans{u[e/x]} \neq False) \cup \etransf{e} \cup \etransf{u[e/x]}$$

@@

We then want to check (with equinox) that:
 $$\trans{C} \cup \{ \dtrans{f~x_1 \dots x_n = e[f^\star/f]}, \dtrans{f^\star~x_1 \dots x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c} \} \models \bigwedge_{c \in C_f} \strans{f \in c}$$

Which we rewrite as:
 $$\trans{C} \cup \{ \dtrans{f~x_1 \dots x_n = e[f^\star/f]}, \dtrans{f^\star~x_1 \dots x_n = e[f^\star/f]}, \bigwedge_{c \in C_f} \strans{f^\star \in c},  \bigvee_{c \in C_f} \lnot \strans{f \in c} \} \models \bot$$
